Local wind regime induced by giant linear dunes

### 2022-01-29 Paper/main.tex

\usepackage[dvipsnames]{xcolor}
\newcommand{\cg}{\textcolor{WildStrawberry}}

\begin{document}

\title{Local wind regime induced by the presence of giant linear dunes.}

\author{Cyril Gadal \and Pauline Delorme \and Clément Narteau \and
        Giles Wiggs \and Matthew Baddock \and Joanna M. Nield \and Philippe Claudin }

\institute{C. Gadal \at
              Institut de Mécanique des Fluides de Toulouse (IMFT), Université de Toulouse, CNRS, INPT, UPS, Toulouse, France \\
              \email{cyril.gadal@imft.fr}
           \and
           S. Author \at
              second address
            \and
            T. Author \at
            third address
}

@@

\begin{abstract}

  \keywords{Boundary layer \and Turbulent flow \and Sand dunes \and Fluide-structures interactions}
\end{abstract}

@@

\section{Introduction}

As a flow encounters an obstacle, different interactions can arise depending on the different time and length scales involved. In the case of atmospheric flows, this mainly depends on the part of the atmosphere, schematically composed of a turbulent boundary layer topped by a stably stratified one, with which the obstacle interacts \citep{Stull1988}.
%
At the largest scale, the feedback of mountains on the stratified part results in wave generation as well as significant wind disturbances, such as downslope winds in the lee side \citep{Duran1990}. Inside the boundary layer, the interaction between a turbulent flow and hilly surfaces is for example key to the understanding ocean surface wind-driven waves~\citep{Sullivan2010}, or eolian bedforms in desert \citep{Courrech2015}.

Looking at the flow close to the surface, two different components of the topography feedback can be isolated. First, the flow accelerates on the upwind slope, and slows down on the downwind one, with a maximum velocity slightly upwind of the crest~\citep{Jackson1975, Sykes1980, Hunt1988}. Several measurements are available, performed in subaqueous flumes~\citep{Zilker1977, Zilker1979, Frederick1988} and in eolian field conditions~\citep{Claudin2013, Lu2021}. Second, deflection occurs when the incident flow direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime~\citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of larger aspect ratios obstacles, such as mountain ranges~\citep{Kim2000, Lewis2008, Fernando2019} or sand dunes~\citep{Walker2009, Walker2013, Hesp2015, Smith2017}.

In the case of aeolian bedforms, the measurements mentioned above were made on dunes a few meters high only. The resulting wind disturbances have then been shown to significantly impact the sediment pathways of the system \citep{Hesp2015}, but also to affect the collective behaviour of dune populations trough long-range interactions \citep{Bacik2020}.
%
However, eolian dunes coexists in sand seas at different scales, each representative of their different evolution stages, from emergence at twenty meters wavelengths to giant size, corresponding to kilometric wavelengths and heights of several dozen meters \citep{McKee1979}. Quantifying these two effects for giant dunes is then all the more important for several reasons.
%
First, larger wind disturbances are expected from larger obstacles.
%
Second, the flow--structure coupling is not straightforward due to vertical structure of the atmosphere with which the giant dunes interact due to their sizes \citep{andreotti2009}.
%
Finally, the morphodynamics of aeolian bedforms are strongly dependent on the wind regime~\citep{Livingstone1996}. Thus, small-scale bedforms are sensitive to wind disturbances induced by larger-scale bedforms. This has been demonstrated for impact ripples on dunes \citep{Howard1977,Hood2021}, but has never been studied for medium-scale dunes near giant dunes.

In the literature, arid areas have been studied at the desert scale from climate reanalyses based on global atmospheric models \citep{Blumberg1996, Livingstone2010, Ashkenazy2012, Jolivet2021, Hu2021}, such as ERA-40, ERA-Interim or ERA-5 \citep{Uppala2005, Dee2011, Hersbach2020}. However, the spatial resolution (tens of kilometers) of these reanalyses implies average quantities that do not resolve the smaller scales, ranging from the study of individual dunes to the border of mountains \citep{Livingstone2010}. Lately, the release of ERA5-Land allow to push back this limitation by providing up to 70 years of hourly wind predictions at a 9~km spatial resolution \citep{munoz2021}. However, due to its recent nature, the applicability limitations of this dataset remain to be studied.

Here, we compare local wind speeds and directions measured in four different places across the Namib desert to the regional predictions of the ERA5-Land climate reanalysis. When the measurement stations are surrounded by a relatively flat environment, we show that local measurements and regional predictions agree with each other. On the contrary, in the sand sea, we show that they differ due to the presence of giant dunes. Furthermore, we link the magnitude of these discrepancies to the circadian cycle of the atmospheric boundary layer. Finally, we draw implications for smaller-scale eolian bedforms.

\section{Wind regimes across the Namib Sand Sea}

  We focus on four places across and nearby the Namib desert, highlighting different environments (see Fig.~\ref{Fig1}).
  %
  The Adamax station is located near the Adamax salt pan, in a highly vegetated area. The Huab station, on the coast at the outlet of the Huab river, is in an arid environment exhibiting $60$-m scale barchan dunes. While the surroundings of these two stations are relatively flat, this is not the case for the Deep Sea and South Namib stations. Both are in the interdune between tens of meters high giant linear dunes with kilometric wavelengths and superimposed patterns. In this section, we describe and compare winds from local measurements and climate reanalysis predictions.

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure1.pdf}
    \caption{Wind data used in this study \textbf{a}: Location of the studied sites. \textbf{b--e}: Satellite images of the studied sites (Google-Earth, Maxar Technologies, CNES/Airbus). In each subplot, the left and right wind roses represent the data from the ERA5Land climate reanalysis and the local wind stations, respectively. Note that the bars show the direction towards which the wind blows. The black dots show the location of local wind stations.}
    \label{Fig1}
  \end{figure}

  \subsection{Datasets}

  Local winds are provided by measurement stations located in the four different places (see black dots in Fig.~\ref{Fig1}). The wind strength and direction are sampled every 10 minutes by cup anemometers and wind vanes, at heights between $2$~m and $3$~m depending on the station. The available period of measurements ranges from 1 to 5 discontinuous years distributed between 2012 and 2020 (see Fig.~\ref{Fig1_supp}). We checked that at least one complete seasonal cycle is available at each station.
  %
  Regional winds are extracted at the same locations and periods from the ERA5-Land dataset, which is a replay at a smaller spatial resolution of ERA5, the latest climate reanalysis from the ECMWWF \citep{Hersbach2020, munoz2021}. It provides hourly predictions of the 10-m wind velocity and direction at a spatial resolution of $\sim 9~\textup{km}$ ($0.1^\circ\times0.1^\circ$).

  For comparison, the local measurements are averaged into 1-hr bins centered on the temporal scale of the ERA5-Land estimates (see Fig.~\ref{Fig2_supp}). As the wind velocities of both datasets are provided at different heights, we convert them into shear velocities (see SI section 1), characteristic of the turbulent wind profile, which are then used together with the wind direction for further analysis. The resulting wind data are shown on the wind roses of Fig.~\ref{Fig1}(b--e).

  Finally, the dune properties are computed using autocorrelation on the 30-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007}. For the South Namib and Deep Sea stations, we obtain respectively orientations of $85^\circ$ and $125^\circ$, wavelengths of $2.6~\textup{km}$ and $2.3~\textup{km}$ and amplitudes of $45~\textup{m}$ and $20~\textup{m}$ (see Fig.~\ref{Fig4_supp} for more details).

  \subsection{Agreement between local and regional winds}

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure2.pdf}
    \caption{Temporal comparison between the wind data coming from the Era5Land climate reanalysis (orange lines) and from the local measurements (blue lines). Color swathes indicate day (between 1000 UTC and 2200 UTC) and night (before 1000 UTC or after 2200 UTC) \textbf{a--b}: Huab station. \textbf{c--d}: Deep Sea station in winter. \textbf{e--f}: Deep Sea station in summer.}
    \label{Fig2}
  \end{figure}

  The obtained wind regimes are shown in figure~\ref{Fig1}. In the Namib, the regional wind patterns are essentially controlled by the see breeze, resulting in strong northward components (sometimes slightly deviated by the large scale topography) present in all regional wind roses \citep{lancaster1985}. These daytime winds are dominant during the second-half of the year (Septembre-January). In winter, an additional easterly component can be recorded during the night, induced by the combination of katabatic winds forming on the mountains, and infrequent `berg' winds, which are responsible of the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from the inland to the coast, resulting in bidirectional wind regimes within the Namib Sand Sea and at the Adamax salt pan (Fig.~\ref{Fig1}b, \ref{Fig1}d and \ref{Fig1}e) and a unidirectional wind regime on the coast at the outlet of the Huab River (Fig.~\ref{Fig1}c).

  In the case of the Adamax and Huab stations, the wind roses from the regional predictions qualitatively match those corresponding to the local measurements. However, for the Deep Sea and South Namib stations, the measured wind roses exhibit additional components aligned with the giant dune orientation visible on the satellite images (Fig.~\ref{Fig1}c--d).
  %
  The time series of wind speed and direction show that this agreement in the case of Adamax and Huab stations is always verified (Fig.~\ref{Fig2}a--b) and Fig.~\ref{Fig5_supp}). In contrast, for the stations within the giant dune field, we observe that this agreement is limited to Septembre--January time periods (Fig.~\ref{Fig2}c--d).

  \subsection{Influence of the giant dunes on local wind regimes}
  \label{section_data_feedback}

  In the February--August period, when giant dunes are present, the local and regional winds match during daytime only, i.e when the southerly/southwesterly sea breeze dominates (see Fig.~\ref{Fig2}(e--f), Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}). In the late afternoon and during the night, when the northwesterly `berg' and katabatic winds blow, measurements and predictions differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes separated of $\simeq 180^\circ$, each corresponding to the giant dune alignement (see the purple frame in Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}, as well as Fig.~\ref{Fig7_supp}). This deviation is also associated with a global attenuation of the wind strength (Fig.~\ref{Fig8_supp}). Remarkably, all these figures show that this process occurs for low wind velocities, typically for $u_{*} < 0.1~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $0.25~\textrm{m}~\textrm{s}^{-1}$, this wind reorientation does not occur. Finally, for intermediate shear velocities, both reorientation along the dune crest and no reorientation are observed (Fig.~\ref{Fig7_supp}).

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure3.pdf}
    \caption{Distributions of wind direction at the Deep Sea Station for the Era5Land climate reanalysis (orange) and the local measurements (blue). In each subplot, both distributions are plotted from the same time steps, selected with constraints on the wind direction (columns) and/or wind velocity (rows) of the Era5Land dataset. The gray vertical dashed lines indicate the dune orientation. The numbers at the top right give the percentage of time steps selected in each sub-range, as well as the percentage of them corresponding to the day (between 1000 UTC and 2200 UTC). The purple frame highlights the regime (small wind velocities, nocturnal summer wind) in which the data from both datasets differs. A similar figure can be obtained for the Deep Sea station (see Fig.~\ref{Fig6_supp}).}
    \label{Fig3}
  \end{figure}

  \section{Influence of the circadian cycle of the atmospheric boundary layer}

  For linear ridges, dune-induced flow disturbances have mainly been related to the angle between wind direction and crest orientation, with a maximum for angles between $30^{\circ}$ and $70^{\circ}$~\citep{Walker2009, Hesp2015}. In our case, the most deflected wind for both stations is the most perpendicular, such that the incident wind direction does not seem to be the dominant parameter controlling the wind deflection. In contrast, a different behavior is observed between  low and high wind velocities, suggesting a change in hydrodynamical regime.

  In the following, we discuss the relevant parameters leading to different hydrodynamical interactions with topographical obstacles, and interpret the data with respect to the corresponding physical mechanisms.

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure4.pdf}
    \caption{\textbf{a--b}: Vertical profiles of the virtual potential temperature at 2 different time steps (day - 31/03/2017 - 1200 UTC, night - 21/03/2017 - 2200 UTC) at the Deep Sea station. Dots: data from the ERA5 reanalysis. Transparent dashed lines: boundary layer height given by the ERA5 reanalysis, calculated from the bulk Richardson number \citep{seidel2012}. Plain lines: vertical (boundary layer) and linear (free atmosphere) fits to estimate the stratification properties. \textbf{c--f}: Sketches representing the interaction between the giant dunes and the atmospheric flow for different meteorological conditions. \textbf{g}: Streamlines qualitatively representing the effect of low, medium and high flow confinement. For details on the streamline derivation, see Appendix~\ref{turbulent_wind_model}.}
    \label{Fig4}
  \end{figure}

  \subsection{Relevant non-dimensional parameters and physical considerations}
  \label{theoretical_framework}

  Flow deflection over ridges can be understood from the Bernoulli principle~\citep{Hesp2015}. As the flow approaches the ridge crest, the compression of the streamlines results in larger flow velocities, and thus lower pressures~\citep{Rubin1987}. An incident flow oblique to the ridge is then deflected towards lower pressure zones, i.e towards the crest. Turbulent dissipation at the bottom and non-linearities tends to increase this effect downstream, resulting in along the crest wind deflection in the lee side~\citep{Hesp2015, Gadal2019}.

  Another way to increase the flow deflection is its confinement below a capping surface, that results in further streamline compression. This is the case for bedforms forming in open channel flows such as rivers~\citep{Fourriere2010, Unsworth2018}, but also for eolian dunes. These dunes evolve in the turbulent atmospheric boundary layer (ABL), which is capped by a transitional layer separating it from the stratified atmosphere above (see Fig.~\ref{Fig4}). Two different mechanisms control the possibility of this additional streamline compression.

  On one hand, it depends if the flow disturbance induced by the underlying topography reach the surface. As obstacles typically disturb flow over a characteristic height similar to their length, the potential of interaction between the dunes and the overlying surface is well captured by the parameter $k H$, where $k = 2\pi/\lambda$ is the wavenumber and $H$ the ABL depth. Note that $H$ is directly related to the radiative fluxes at the Earth surface, and thus varies with the circadian and seasonal cycles. Here, the giant dunes have kilometric wavelengths, such that $0.02 \lesssim k H \lesssim 5$, and they interact most of the time with the capping layer and the stratified free atmosphere (FA) above \citep{andreotti2009}. Interestingly, the limit of no-interactions between the topography and the boundary layer structure ($k H \gg 1$), in which the properties of the overlying atmospheric structure are irrelevant, is never reached here, in the case of giant dunes.

  On the other hand, it depends on rigidity of the capping surface, as its deformation releases the confinement effect inducing streamline compression. This is typically quantified using the Froude number~\citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014}:
  \begin{equation}
        \mathcal{F} = \displaystyle\frac{U}{\sqrt{\displaystyle\frac{\Delta\rho}{\rho}gH}},
  \end{equation}
  where $U$ is the wind velocity at the top of the ABL, $\rho$ its average density, $\Delta\rho$ the density jump between the ABL and the FA.

  % Note that the ability of the capping layer and stratification to accommodate a perturbation induced by the topography directly impacts the strength of this confinement effect (Fig.~\ref{Fig4}). This is typically quantified using surface and internal Froude numbers
  % \citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014}:
  % \begin{equation}
  %       \mathcal{F} = \displaystyle\frac{U}{\sqrt{\displaystyle\frac{\Delta\rho}{\rho}gH}}, \, \mathcal{F}_{\textup{I}} = \displaystyle\frac{k U}{N},
  % \end{equation}
  % where $U$ is the wind velocity at the top of the ABL, $\rho$ its average density, $\Delta\rho$ the density jump between the ABL and the FA and $N$ is the Brunt-Väisälä frequency, characteristic of the stratification.

  The smallest wind disturbances are expected during the day, when the ABL depth is comparable to the dune wavelength ($k H \gtrsim 1$) and for large wind velocities, which correspond to a weak confinement situation (Fig.~\ref{Fig4}d). On the contrary, large wind disturbances are expected to occur during the night, when the confinement is mainly induced by shallow ABL (Fig.~\ref{Fig4}e--f). Note that this strong confinement can be somewhat reduced in the case of strong winds (corresponding to large Froude numbers, see Fig.~\ref{Fig4}f), explaining the transition from deflected to non-deflected winds related to low and high velocities observed in the data (see section~\ref{section_data_feedback}).

  \subsection{Flow regime diagrams}

>>>>

\usepackage[dvipsnames]{xcolor}
\newcommand{\cg}{\textcolor{WildStrawberry}}
\newcommand{\pc}{\textcolor{ForestGreen}}
\newcommand{\new}{\textcolor{blue}}
\newcommand{\todo}{\textcolor{red}}

%---------------------------------------------------------------------------------------
\begin{document}

\title{Local wind regime induced by giant linear dunes}

\author{Cyril Gadal \and Pauline Delorme \and Cl\'ement Narteau \and Giles F.S. Wiggs \and Matthew Baddock \and Joanna M. Nield \and Philippe Claudin}

\institute{
C. Gadal \at
Institut de M\'ecanique des Fluides de Toulouse, Universit\'e de Toulouse Paul Sabatier, CNRS, Toulouse INP-ENSEEIHT, Toulouse, France. \email{cyril.gadal@imft.fr}
\and
P. Delorme \at
School of Geography and Environmental Science, University of Southampton, Southampton, UK.
\and
C. Narteau \at
Institut de Physique du Globe de Paris, Universit\'e de Paris, CNRS, Paris, France.
\and
G.S.F. Wiggs \at
School of Geography and the Environment, University of Oxford, Oxford, UK.
\and
M. Baddock \at
Geography and Environment, Loughborough University of Technology, Loughborough, UK.
\and
J.M. Nield \at
School of Geography and Environmental Science, University of Southampton, Southampton, UK.
\and
P. Claudin \at
Physique et M\'ecanique des Milieux H\'et\'erog\`enes, CNRS, ESPCI Paris, PSL Research University, Universit\'e de Paris, Sorbonne Universit\'e, Paris, France.  \email{philippe.claudin@espci.fr}
}

@@

\begin{abstract}
Emergence and growth of sand dunes result from the interaction between the topography, which perturbs the wind flow, which itself controls the sediment transport. While these feedbacks are well studied at the scale of a single dune, the average effect of a periodic dune pattern on atmospheric flows remains poorly constrained due to a lack of data in major sand seas. Here, we compare field measurements of surface wind data to the predictions of the ERA5-Land climate reanalysis in four different places in Namibia, inside and north of the giant-dune field of the Namib sand sea. In the flat desert areas to the north, observations and predictions agree well despite the small amount of data in the assimilation process. This is also the case in the interdune areas of the sand sea, except for the weak winds blowing at night in a direction oblique to the orientation of the giant dunes. We quantify these similarities and differences and provide a physical understanding of the relevant aerodynamical regimes to relate them to the daily cycle of the turbulent atmospheric boundary layer over a dune pattern of given wavelength. We conclude by identifying the conditions under which the ERA5-Land reanalysis data can or can't be used to study dune morphodynamics. We also propose that, in multidirectional wind regimes, deflections of specific winds could explain the occurrence of secondary dune patterns with a different orientation than the primary structures between which they grow.
\keywords{Atmospheric boundary layer \and Sand dunes \and Fluid-structure interactions}
\end{abstract}

@@

%---------------------------------------------------------------------------------------
\section{Introduction}

The description of turbulent flows over a complex topography is a rich and active subject, that is relevant for a large variety of different environmental systems \citep{Finnigan2020}. For example, the flow over hills is of primary interest for wind power, meteorological and air pollution phenomena \citep{Taylor1987}. The properties of these flows are also key to the understanding of the formation of ocean surface wind-driven waves \citep{Sullivan2010}, dissolution bedforms \citep{Claudin2017}, or sedimentary ripples and dunes \citep{Charru2013, Courrech2015}. Importantly, the troposphere presents a vertical structure, with a lower convective boundary layer, of typical kilometer-scale thickness, capped by a stably stratified region \citep{Stull1988}. The largest topographic obstacles as mountains can therefore interact with this upper region, for example leading to internal wave generation or significant wind disturbances, such as downslope winds in lee sides \citep{Durran1990}.

Looking at the wind close to the surface, two topographic feedbacks on the flow, although related, can be commented separately. There is a first effect on the wind intensity: the flow accelerates on the upwind slope and slows down on the downwind one \citep{Baddock2011}, with a speed-up factor essentially proportional to the obstacle aspect ratio \pc{here a ref?}. Importantly, the velocity maximum is typically shifted upwind of the obstacle crest. This behaviour has been theoretically predicted by means of asymptotic analysis of a neutrally stratified boundary-layer flow over an obstacle of vanishing aspect ratio \citep{Jackson1975, Mason1979, Sykes1980, Hunt1988, Belcher1998}. Experiments in flumes \citep{Zilker1977, Zilker1979, Frederick1988, Poggi2007}, in wind tunnel \citep{Gong1989, Finnigan1990, Gong1996} and in field conditions \citep{Taylor1987a, Claudin2013, Fernando2019, Lu2021}, have also documented this effect. Interestingly, a similar behaviour exists for the pressure perturbation, but with a slight downwind shift for the pressure minimum \citep{Claudin2021}. The second effect is the flow deflection that occurs when the incident wind direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime valid for shallow topography \citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of steep enough obstacles, such as mountain ranges \citep{Kim2000, Lewis2008, Fernando2019} or well developed sand dunes \citep{Walker2009, Hesp2015, Walker2017, Smith2017}.

For practical reasons, wind measurement over sand dunes have been performed over rather small bedforms, typically a few meters high only -- that is tens of meters in length. Giant dunes, with kilometer-scale wavelengths and heights of tens of meters, are more difficult to investigate although they provide, for several reasons, a choice configuration for the study of turbulent flows over a complex topography. First of all, one expects larger wind disturbances for larger obstacles. Second, their large size make them interact with the vertical structure of the atmosphere \citep{andreotti2009}. Also, they usually form large patterns in sand seas and thus behave as rather clean periodic perturbations, in contrast with isolated hills. Finally, because the morphodynamics of aeolian bedforms are strongly dependent on the local wind regime \citep{Livingstone1996}, one can expect to see the consequences of the wind disturbances on smaller dunes, in a similar way as what has been reported for the effect of dunes on impact ripples \citep{Howard1977,Hood2021}.

Arid areas have been much studied at the desert scale from climate reanalyses based on global atmospheric models \citep{Blumberg1996, Livingstone2010, Ashkenazy2012, Jolivet2021, Hu2021}, such as ERA-40, ERA-Interim or ERA-5 \citep{Uppala2005, Dee2011, Hersbach2020}. However, the spatial resolution (tens of kilometers) of these reanalyses implies average quantities that do not resolve the smaller scales, ranging from individual dunes to the border of mountains \citep{Livingstone2010}. Lately, the release of ERA5-Land allows to push back this limitation by providing up to 70 years of hourly wind predictions at a $9$~km spatial resolution \citep{munoz2021}. However, its validity remains to be studied, especially in areas where data assimilation is low (such as deserts).

In this work, we compare local wind speeds and directions measured in four different places inside and north of the giant-dune field of the Namib sand sea to the regional predictions of the ERA5-Land climate reanalysis. When the meteorological stations are surrounded by a relatively flat environment, we show that local measurements and regional predictions agree well with each other. The find that the agreement is also good in the interdune areas of the sand sea, except for the weak winds blowing at night in a direction oblique to the orientation of the giant dunes (section 2). Furthermore, we are able to link the magnitude of these differences to the circadian cycle of the atmospheric boundary layer (section 3). Finally, we draw implications of the wind disturbances on smaller-scale dunes (section 4).

%---------------------------------------------
\begin{figure}[t]
\centering
\includegraphics[scale=1]{Figures/Figure1.pdf}
\caption{Wind data used in this study \textbf{a}: Location of the different sites in Namibia. \textbf{b-e}: Satellite images of these different environements (Google-Earth, Maxar Technologies, CNES/Airbus). In each subplot, the left and right wind roses represent the data from the ERA5-Land climate reanalysis and the local wind stations, respectively. Note that the bars show the direction towards which the wind blows. The black dots show the location of local wind stations.}
\label{Fig1}
\end{figure}
%---------------------------------------------

%---------------------------------------------------------------------------------------
\section{Wind regimes across the Namib Sand Sea}
We have measured the wind regime in four different sites of Namibia, representative of various environments across and nearby the Namib desert (Fig.~\ref{Fig1}). The Adamax station is located near the Adamax salt pan, in a highly vegetated area. The Huab station, on the coast at the outlet of the Huab river, is in an arid place where $60$-m scale barchan dunes develop. These two stations are located in a relatively flat \pc{quantify how flat is flat ?} environment. In contrast, the Deep Sea and South Namib stations are located in the interdune between tens of meters high giant linear dunes with kilometer-scale wavelengths and superimposed patterns. In this section, we describe and compare winds from local measurements and climate reanalysis predictions.

%------------------------
\subsection{Wind and elevation data}
Local winds are measured at meteorological stations located in these four different places (black dots in Fig.~\ref{Fig1}). The wind strength and direction are sampled every 10 minutes by cup anemometers and wind vanes, at heights between $2$~m and $3$~m depending on the station. The available period of measurements ranges from 1 to 5 discontinuous years distributed between 2012 and 2020 (Suppl. Fig.~\ref{Fig1_supp}). We checked that at least one complete seasonal cycle is available at each station. Regional winds are extracted at the same locations and periods from the ERA5-Land dataset, which is a replay at a smaller spatial resolution of ERA5, the latest climate reanalysis from the ECMWWF \citep{Hersbach2020, munoz2021}. It provides hourly predictions of the 10-m wind velocity and direction at a spatial resolution of $0.1^\circ\times0.1^\circ$ ($\simeq 9~\textup{km}$ in Namibia).

For comparison, the local measurements are averaged into 1-hr bins centered on the temporal scale of the ERA5-Land estimates (Suppl. Fig.~\ref{Fig2_supp}). As the wind velocities of both datasets are provided at different heights, we convert them into shear velocities (Suppl. Mat. section 1), characteristic of the turbulent wind profile, which are then used together with the wind direction for further analysis. Wind roses in Fig.~\ref{Fig1}(b--e) show the resulting wind data.

Dune properties are computed using autocorrelation on the 30-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007}. For the South Namib and Deep Sea stations, we obtain respectively orientations of $85^\circ$ and $125^\circ$ with respect to the North, wavelengths of $2.6~\textup{km}$ and $2.3~\textup{km}$ and amplitudes of $45~\textup{m}$ and $20~\textup{m}$ (Suppl. Fig.~\ref{Fig4_supp} for more details).

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure2.pdf}
\caption{Temporal comparison between the wind data coming from the ERA5-Land climate reanalysis (orange lines) and from the local measurements (blue lines). Coloured swathes indicate day (between 1000 UTC and 2200 UTC) and night (before 1000 UTC or after 2200 UTC) \textbf{a-b}: Huab station. \textbf{c-d}: Deep Sea station in winter. \textbf{e-f}: Deep Sea station in summer.}
\label{Fig2}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Comparison of local and regional winds}
\label{section_data_feedback}
The obtained wind regimes are shown in figure~\ref{Fig1}. In the Namib, the regional wind patterns are essentially controlled by the see breeze, resulting in strong northward components (sometimes slightly deviated by the large scale topography \pc{mountain?}) present in all regional wind roses \citep{lancaster1985}. These daytime winds are dominant during the second-half of the year (September-January). In winter, an additional easterly component can be recorded during the night, induced by the combination of katabatic winds forming in the mountains, and infrequent `berg' winds, which are responsible of the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from the inland to the coast. As a result, bidirectional wind regimes within the Namib Sand Sea and at the Adamax salt pan (Fig.~\ref{Fig1}b,d,e) and a unidirectional wind regime on the coast at the outlet of the Huab River (Fig.~\ref{Fig1}c) are observed.

In the case of the Adamax and Huab stations, the time series of wind speed and direction from the regional predictions quantitatively match those corresponding to the local measurements (Fig.~\ref{Fig2}a,b) and Suppl. Fig.~\ref{Fig5_supp}). For the Deep Sea and South Namib stations within the giant dune field, we observe that this agreement is also good, but limited to the September-January time period (Fig.~\ref{Fig2}c,d). As a matter of fact, the measured wind roses exhibit additional wind components aligned with the giant dune orientation, as evidenced on the satellite images (Fig.~\ref{Fig1}c,d).

More precisely, during the February-August period, the local and regional winds in the interdune match during daytime only, i.e when the southerly/southwesterly sea breeze dominates (Figs.~\ref{Fig2}(e,f) and \ref{Fig3}, Suppl. Fig.~\ref{Fig6_supp}). In the late afternoon and during the night, when the northwesterly `berg' and katabatic winds blow, measurements and predictions differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes corresponding to reversing winds aligned with the giant dune orientation (purple frame in Fig.~\ref{Fig3}, Suppl. Figs.~\ref{Fig6_supp} and \ref{Fig7_supp}). This deviation is also associated with a global attenuation of the wind strength (Suppl. Fig.~\ref{Fig8_supp}). Remarkably, all these figures show that this wind reorientation and attenuation processes typically occur at low wind velocities, typically for $u_{*} \lesssim 0.1~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $\simeq 0.25~\textrm{m}~\textrm{s}^{-1}$, the wind reorientation is not significant. Finally, for intermediate shear velocities, moments of reorientation along the dune crest and others with no reorientation can both be observed (Suppl. Fig.~\ref{Fig7_supp}). \pc{CHECK the above lines cf. $\delta u$}

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure3.pdf}
\caption{Distributions of wind direction at the Deep Sea Station for the ERA5-Land climate reanalysis (orange) and the local measurements (blue). In each subplot, both distributions are plotted from the same time steps, selected with constraints on the wind direction (columns) and/or wind velocity (rows) of the ERA5-Land dataset. The gray vertical dashed lines indicate the dune orientation. The numbers at the top right give the percentage of time steps selected in each sub-range, as well as the percentage of them corresponding to the day (between 1000 UTC and 2200 UTC). The purple frame highlights the regime (small wind velocities, nocturnal summer wind) in which the data from both datasets differ. A similar figure can be obtained for the Deep Sea station (Suppl. Fig.~\ref{Fig6_supp}).}
\label{Fig3}
\end{figure}
%---------------------------------------------

%---------------------------------------------------------------------------------------
\section{Influence of wind speed and circadian cycle on the atmospheric boundary layer}
The wind deflection induced by linear dunes has so far mainly been related to the angle between wind direction and crest orientation, with a maximum deflection for angles between $30^{\circ}$ and $70^{\circ}$ \citep{Walker2009, Hesp2015}. In the analysed data, the most deflected wind at both Deep Sean and South Namib stations is the most perpendicular to the giant dunes (\pc{Fig.~\ref{***}}). The incident wind direction then does not seem to be the only parameter controlling the wind deflection. In contrast, a different behaviour is observed between low and high wind velocities, suggesting a change in hydrodynamical regime. In this section, we discuss the relevant parameters associated with the dynamical mechanisms that govern the interactions between the atmospheric boundary layer flow and topographical obstacles. This analysis allows us to provide a physics-based interpretation the observations.

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure4.pdf}
\caption{\textbf{a-b}: Vertical profiles of the virtual potential temperature at 2 different time steps (day - 03/11/2015 - 1200 UTC, night - 01/13/2013 - 0900 UTC) at the Deep Sea station. Dots: data from the ERA5 reanalysis. Dashed lines: boundary layer height given by the ERA5 reanalysis (Suppl. Mat. section 2). Plain lines: vertical (boundary layer) and linear (free atmosphere) fits to estimate the stratification properties. \textbf{c-f}: Sketches representing the interaction between the giant dunes and the atmospheric flow for different meteorological conditions. \textbf{g}: Streamlines over a sinusoidal topography $\xi(x,y)$ qualitatively representing the effect of low, medium and strong flow confinement, in relation to the above panels (see Appendix~1 for more details). \pc{add arrow with reference wind}}
\label{Fig4}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Flow over a modulated bed}
\label{theoretical_framework}
Taking as a reference the turbulent flow over a flat bed, the general framework of this study is the understanding and the description of the flow response to a bed modulation. Without loss of generality, we can consider in this context an idealised bed elevation in the form of parallel sinusoidal ridges, with wavelength $\lambda$ (or wavenumber $k = 2\pi/\lambda$) and amplitude $\xi_0$, and where the reference flow direction makes a given angle with respect to the ridge crest \citep{Andreotti2012}. Part of this response, on which we focus here, is the flow deflection by the ridges. In a simplified way, it can be understood from the Bernoulli principle~\citep{Hesp2015}: as the flow approaches the ridge crest, the compression of the streamlines results in larger flow velocities, and thus lower pressures \citep{Rubin1987}. An incident flow oblique to the ridge is then deflected towards lower pressure zones, i.e towards the crest. Turbulent dissipation tends to increase this effect downstream, resulting in along the crest wind deflection in the lee side~\citep{Gadal2019}.
